---
title: "Biostat 203B Homework 1"
author: Lillian Chen
date: Due Jan 22 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval=F}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email.

**Solution**: Done.

2. Create a **private** repository `biostat-203b-2021-winter` and add `Hua-Zhou`, `Chris-German` and `ElvisCuiHan` as your collaborators with write permission.

**Solution**: Done.

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `master` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `master` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and data sets to reproduce results) in `master` branch.

**Solution**: Done.

4. After each homework due date, teaching assistant and instructor will check out your master branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

**Solution**: Done.

5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

**Solution**: Done.

## Q2. Linux Shell Commands

1. This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic-iv.mit.edu/docs/access/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)

[Completion Report Link](https://www.citiprogram.org/verify/?k7c4d75ad-018d-4743-9ade-2a32928f11ca-40330161)

[Completion Certificate Link](https://www.citiprogram.org/verify/?wd16c55a9-7700-4c14-9850-604284d721ef-40330161)

2. The `/usr/203b-data/mimic-iv/` folder on teaching server contains data sets from MIMIC-IV. Refer to <https://mimic-iv.mit.edu/docs/datasets/> for details of data files.  
    ```{bash}
    ls -l /usr/203b-data/mimic-iv
    ```
Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/usr/203b-data/mimic-iv` directly in following exercises. 

    Use Bash commands to answer following questions.

3. Display the contents in the folders `core`, `hosp`, `icu`. What are the functionalities of the bash commands `zcat`, `zless`, `zmore`, and `zgrep`? 

```{bash}
echo "Contents of 'core'"
ls -l /usr/203b-data/mimic-iv/core/
echo '' # line break
echo "Contents of 'hosp'"
ls -l /usr/203b-data/mimic-iv/hosp/
echo '' # line break
echo "Contents of 'icu'"
ls -l /usr/203b-data/mimic-iv/icu/
```

`zcat` - compresses or expands files

`zless` - file perusal filter for crt viewing of compressed text one screenful at a time on a soft-copy terminal. It has more functionalities than `zmore` and loads files faster than `zmore`.

`zmore` - file perusal filter for crt viewing of compressed text, primary function is viewing compressed text by scrolling downward.

`zgrep` - search possibly compressed files for lines that match a regular expression


4. What's the output of following bash script?
    ```{bash, eval=F}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        ls -l $datafile
      done
    ```
Display the number of lines in each data file using a similar loop.

**Solution**: 

The bash script above lists the datafiles of .gz type within the core directory found with path /usr/203b-data/mimic-iv/core/.

There are 524,521 lines in `admissions.csv.gz`, 383,221 lines in `patients.csv.gz`, and 2,192,964 lines in `transfers.csv.gz`.
```{bash}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        zcat $datafile | wc -l
      done
```



5. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, `language`, `marital_status`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq`, `wc`, and so on.)

**Solution**:


Here are the first few (10) rows of `admissions.csv.gz`:
```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | head
```

There are 524521 rows in data file 'admissions.csv.gz'.
```{bash, eval=F}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | wc -l
```

There are 257366 unique patients identified by `subject_id` in this data file.
```{bash, eval=F}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |  
awk -F, '{if (NR>1) print $1}' | sort | uniq | wc -l
```

There are 9 possible values taken by variable `admission_type`. Counts are reported in parentheses after each possible value. 

AMBULATORY OBSERVATION (7259), DIRECT EMER. (21595), DIRECT OBSERVATION (20014), EU OBSERVATION (100938), EW EMER. (158032), OBSERVATION ADMIT (55532), SURGICAL SAME DAY ADMISSION (41103), URGENT (47970). 

```{bash, eval=F}
# possible values for admission_type
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | 
awk -F, '{if (NR>1) print $6}' | sort | uniq -c

# number of possible values taken
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | 
awk -F, '{if (NR>1) print $6}' | sort | uniq | wc -l
```

There are 12 possible values taken by variable `admission_location`. Counts are reported in parentheses after each possible value.

' ' (blank) (60437), AMBULATORY SURGERY TRANSFER (191), CLINIC REFERRAL (10678), EMERGENCY ROOM (246241), INFORMATION NOT AVAILABLE (379), INTERNAL TRANSFER TO OR FROM PSYCH (4484), PACU (6075), PHYSICIAN REFERRAL (127648), PROCEDURE SITE (8453), TRANSFER FROM HOSPITAL (39150), TRANSFER FROM SKILLED NURSING FACILITY (4069), WALK-IN/SELF REFERRAL (16715).
```{bash, eval=F}
# possible values for admission_location
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | 
awk -F, '{if (NR>1) print $7}' | sort | uniq -c

# number of possible values taken
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | 
awk -F, '{if (NR>1) print $7}' | sort | uniq | wc -l
```

There are 3 possible values taken by variable `insurance`. Counts are reported in parentheses after each possible value. 

Medicaid (50962), Medicare (171492), Other (302066).
```{bash, eval=F}
# possible values for insurance
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | 
awk -F, '{if (NR>1) print $9}' | sort | uniq -c

# number of possible values taken
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | 
awk -F, '{if (NR>1) print $9}' | sort | uniq | wc -l 
```

There are 2 possible values taken by variable `insurance`. Counts are reported in parentheses after each possible value.

? (46231), ENGLISH (478289).
```{bash, eval=F}
# possible values for language
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | 
awk -F, '{if (NR>1) print $10}' | sort | uniq -c

# number of possible values taken
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |
awk -F, '{if (NR>1) print $10}' | sort | uniq | wc -l
```


There are 5 possible values taken by variable `insurance`. Counts are reported in parentheses after each possible value.

' ' (blank) (66133), DIVORCED (33762), MARRIED (195326), SINGLE (180453), WIDOWED (48846).
```{bash, eval=F}
# possible values for marital_status
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | 
awk -F, '{if (NR>1) print $11}' | sort | uniq -c

# number of possible values taken
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | 
awk -F, '{if (NR>1) print $11}' | sort | uniq | wc -l
```

There are 8 possible values taken by variable `ethnicity`. Counts are reported in parentheses after each possible value.

AMERICAN INDIAN/ALASKA NATIVE (1536), ASIAN (24522), BLACK/AFRICAN AMERICAN (80526), HISPANIC/LATINO (29887), OTHER (26844), UNABLE TO OBTAIN (3742), UNKNOWN (19419), WHITE (338044).
```{bash, eval=F}
# possible values for ethnicity
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | 
awk -F, '{if (NR>1) print $12}' | sort | uniq -c

# number of possible values taken
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | 
awk -F, '{if (NR>1) print $12}' | sort | uniq | wc -l
```


## Q3. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=FALSE}
    curl http://www.gutenberg.org/cache/epub/42671/pg42671.txt > pride_and_prejudice.txt
    ```
Do **not** put this text file `pride_and_prejudice.txt` in Git. Using a `for` loop, how would you tabulate the number of times each of the four characters is mentioned?


**Solution**:

I used a `for` loop iterating the names into the `grep` command to search for the name and output the count using the `-o` option (`-o` outputs the matches, ignoring lines). `wc -l` counts the number of lines in the output. 'Elizabeth' appeared 634 times, 'Darcy' appeared 417 times, 'Jane' appeared 294 times, and 'Lydia' appeared 170 times.
```{bash, eval=F}
character=('Elizabeth' 'Darcy' 'Jane' 'Lydia')
for name in "${character[@]}"
  do
    echo $name':'
    grep -o $name pride_and_prejudice.txt | wc -l
  done

```


2. What's the difference between the following two commands?
    ```{bash eval=FALSE}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```
**Solution**:

The first command creating test1.txt with the `>` indicator creates or replaces the file each time with a file that contains the text 'hello, world'. The second command for test2.txt with the `>>` indicator creates the file with the 'hello, world' text if it does not exist yet, and appends the 'hello, world' text if the file already exists in the directory.


3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash}
    ./middle.sh pride_and_prejudice.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?


**Solution**:


I ran `chmod +x middle.sh` to make the file executable by the owner. The output prints 5 lines until line 20 from the middle of the text file, which in pride_and_prejudice.txt specifies lines for the author and the editor (lines 16-20). We need the first line of the shell script in order to execute the commands using the `sh` shell, since we are using bash shell in the code chunk above. The exclamation mark in line 1 is formatting related.


### Q4. More fun with Linux

Try these commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.

**Solution**:


`cal` displays the calendar of the current month
```{bash}
cal
```


`cal 2021` displays the monthly calendar of the entire 2021 year
```{bash}
cal 2021
```


`cal 9 1752` tries to display the calendar of September 1752, but there are dates missing (missing days 3-13).
```{bash}
cal 9 1752
```


`date` gives the current date and time in military time using UTC time zone.
```{bash}
date
```


`hostname` prints the name of the host server.
```{bash}
hostname
```


`arch` prints the machine hardware name. For me, it printed 'x86_64'.
```{bash}
arch
```


`uname -a` prints system information, with the `-a` option specifying to print all information
```{bash}
uname -a
```


`uptime` tells you how long the system has been running, and additionally tells you the current time, the number of users logged on, and the system load averages for the past 1, 5, and 15 min.
```{bash}
uptime
```


`whoami` prints the username of the current user running the terminal. For me, it prints 'lchen121'.
```{bash}
whoami
```


`who` prints the identifying information for all users logged onto the server.
```{bash}
who
```


`w` shows the activity of currently logged on users.
```{bash}
w
```


`id` shows all the user and group information for the current user.
```{bash}
id
```


`last` shows the first ten listings of last logged in users.
```{bash}
last | head
```


`echo {con,pre}{sent,fer}{s,ed}` displays text that chooses one option from each of the braces and concatenates them into a string.
```{bash}
echo {con,pre}{sent,fer}{s,ed}
```


`time sleep 5` times the execution of a script, with a 5 second delay specified (sleep delays for the specified amount of time; in this case, 5 (seconds))
```{bash}
time sleep 5
```


`history | tail` prints the last few (up to 10) lines of commands that were entered in the terminal window / in bash. I had to run this with `set -o history` before `history | tail` so that the history recording is enabled.
```{bash}
set -o history
history | tail
```

