---
title: "Biostat 203B Homework 2"
subtitle: Due Feb 5 @ 11:59PM
output: 
  html_document:
    toc: true
    toc_depth: 4 
---

Display machine information for reproducibility:
```{r}
sessionInfo()
```

```{r setup, include = F, eval=T}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(tidyverse)
library(data.table)
library(lubridate)
```

```{r}
os <- sessionInfo()$running
if (str_detect(os, "Linux")) {
  mimic_path <- "/usr/203b-data/mimic-iv"
} else if (str_detect(os, "macOS")) {
  mimic_path <- "/Users/huazhou/Documents/Box Sync/MIMIC/mimic-iv-0.4"
}
```

Use tidyverse (ggpot2, dplyr) to explore the [MIMIC-IV](https://mimic-iv.mit.edu) data introduced in [homework 1](https://ucla-biostat203b-2021winter.github.io/hw/hw1/hw1.html).

```{r}
system(str_c("tree -s -L 2 ", shQuote(mimic_path)), intern = TRUE)
```

## Q1. PhysioNet credential

At this moment, you should already get credentialed on the PhysioNet. Please include a screenshot of your `Data Use Agreement for the MIMIC-IV (v0.4)`.

## Q2. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package. Is there any speed difference?

**Solution**:

Using `read.csv` is extremely slow compared to `read_csv` and `fread`. `read_csv` and `fread` are similar in speed on my computer, with `fread` being a bit faster. 

```{r Q2}

# setting working directory to the core directory in the mimic-iv dataset
setwd(paste0(mimic_path,"/core"))


system.time({ read.csv("admissions.csv.gz") })

system.time({ read_csv("admissions.csv.gz") })

system.time({ fread("admissions.csv.gz") })


```

In this homework, we stick to the tidyverse. 

## Q3. ICU stays

`icustays.csv.gz` (<https://mimic-iv.mit.edu/docs/datasets/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. Summarize following variables using appropriate numerics or graphs:   

- how many unique `stay_id`?  
- how many unique `subject_id`?  
- length of ICU stay  
- first ICU unit  
- last ICU unit  

There are 69,619 unique `stay_id` values and 50,048 unique `subject_id` values.

I created a variable `stay_length` to represent the length of each ICU stay, and found the difference between `outtime` and `intime` in days. An analysis of the distribution of `stay_length` showed that the data has extreme positive skew, so I applied a log transformation to rectify that and used the log transform of `stay_length`, `log.stay_length`, for the subsequent graphical visualization.

```{r Q3 Analysis, include = F, eval = T}
# Zhou prefers graphs

# setting working directory to the icu directory in the mimic-iv dataset
setwd(paste0(mimic_path, "/icu"))


icustays <- read_csv("icustays.csv.gz")


### numerics

# unique stay_id
length(unique(icustays$stay_id))

# unique subject_id
length(unique(icustays$subject_id))


### analysis

## length of ICU stay
q1p3 <- icustays %>% 
  mutate(stay_length = difftime(outtime, intime, units = "days")) %>% 
  mutate(stay_length = as.numeric(stay_length)) %>% 
  mutate(log.stay_length = log(stay_length))

# qqplot of stay_length shows extreme skew
ggplot(data = q1p3) + 
  geom_qq(aes(sample = stay_length)) +
  theme_minimal() +
  labs(x = "theoretical", y = "observed", 
       title = "Q-Q Plot for Length of ICU Stays")

# qqplot of log.stay_length shows better distribution
ggplot(data = q1p3) + 
  geom_qq(aes(sample = log.stay_length)) +
  theme_minimal() +
  labs(x = "theoretical", y = "observed", 
       title = "Q-Q Plot for Log of Length of ICU Stays")

```


```{r Q3 Graphs}
### Graphs

# histogram for log.stay_length  
ggplot(data = q1p3, mapping = aes(log.stay_length)) + 
  geom_histogram(bins = 30) +
  theme_minimal() +
  labs(x = "Log of Length of ICU Stays", y = "Frequency", 
       title = "Histogram of Log of Length of ICU Stays")


## first ICU unit
ggplot(data = q1p3, mapping = aes(first_careunit)) +
  geom_bar(aes(fill = first_careunit)) +
  scale_x_discrete(labels=c("CVICU", "CCU", "MICU", "MICU/SICU", "Medicine", 
                            "Neuro Intermediate", "Neuro Stepdown", 
                            "Neuro SICU", "PACU", "SICU", "TSICU")) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45), 
        legend.position = "bottom") +
    labs(x = "Care Unit", y = "Count",
       title = "First ICU Unit")


## last ICU unit
ggplot(data = q1p3, mapping = aes(last_careunit)) +
  geom_bar(aes(fill = last_careunit)) +
  scale_x_discrete(labels=c("CVICU", "CCU", "MICU", "MICU/SICU", 
                            "Neuro Intermediate", "Neuro Stepdown", 
                            "Neuro SICU", "SICU", "TSICU")) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45), 
        legend.position = "bottom") + 
  labs(x = "Care Unit", y = "Count",
       title = "Last ICU Unit")
```





## Q4. `admission` data

Information of the patients admitted into hospital is available in `ADMISSION.csv.gz`. See <https://mimic-iv.mit.edu/docs/datasets/core/admissions/> for details of each field in this file. Summarize following variables using appropriate graphs. Explain any patterns you observe.   

- admission year  
- admission month  
- admission month day  
- admission week day  
- admission hour (anything unusual?)  
- number of deaths in each year  
- admission type  
- number of admissions per patient  
- admission location  
- discharge location  
- insurance  
- language  
- martial status  
- ethnicity  
- death 

Note it is possible that one patient (uniquely identified by the `SUBJECT_ID`) is admitted into hospital multiple times. When summarizing some demographic information, it makes sense to summarize based on unique patients. 


```{r Q4}
# summarize using best graph/stuff according to data type (continuous vs categorical etc)
```


## Q5. `patient` data

Explore `patients.csv.gz` (<https://mimic-iv.mit.edu/docs/datasets/core/patients/>) and summarize following variables using appropriate numerics and graphs:  

- `gender`  
- `anchor_age` (explain pattern you see)

```{r Q5}

```


## Q6. Lab results

`labevents.csv.gz` (<https://mimic-iv.mit.edu/docs/datasets/hosp/labevents/>) contains all laboratory measurements for patients. 

We are interested in the lab measurements of creatinine, potassium, sodium, chloride, bicarbonate, hematocrit, white blood cell count, glucose, magnesium, calcium, phosphorus, and lactate. Find the `itemid`s of these lab measurements from `d_labitems.csv.gz` and retrieve a subset of `labevents.csv.gz` only containing these items.


```{r Q6}

```

## Q7. Vitals from chartered events

We are interested in the vitals for ICU patients: heart rate, mean and systolic blood pressure (invasive and noninvasive measurements combined), body temperature, SpO2, and respiratory rate.

`chartevents.csv.gz` (<https://mimic-iv.mit.edu/docs/datasets/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `ITEMID` variable indicates a single measurement type in the database. The `VALUE` variable is the value measured for `ITEMID`. 

`d_items.csv.gz` (<https://mimic-iv.mit.edu/docs/datasets/icu/d_items/>) is the dictionary for the `ITEMID` in `CHARTEVENTS.csv`. Find potential values of `ITEMID` that correspond to systolic blood pressure, i.e., `LABEL` contains the string `systolic`. 


```{r Q7}

```

## Q8. Putting things together

Let us create a tibble for all ICU stays, where rows are  

- first ICU stay of each unique patient  
- adults (age at admission > 18)  

and columns contains at least following variables  

- all variables in `icustays.csv.gz`  
- all variables in `admission.csv.gz`  
- first lab measurements during ICU stay  
- first vitals measurement during ICU stay  
- an indicator variable whether the patient died within 30 days of hospital admission  #30 day mortality Y/N variable

```{r Q8}

```


